\ourlongtitle

### 2021-08-23 Bernoulli/glo-2020-numerics.tex
Corrected typos

In the following we will consider an instance of Example~\ref{eq:durham:gallant}, Case~1, where the latent diffusion $(X_t)_{t > 0}$ is an \emph{Ornstein--Uhlenbeck process} \cite{uhlenbeck1930theory} on $\rset$ parameterised by
$$
\mu(x) = - (x - \theta), \quad \sigma(x) \equiv 1,
$$
where $\theta \in \mathbb{R}$ is a parameter.
This process is assumed to be initialised according to the standard normal distribution. Furthermore, conditionally to $(X_t)_{t > 0}$, observations $(Y_n)_{n \in \nset}$ are generated as
$$
Y_n = (1 - \precpar) \varphi(X_n) + \eta_n, \quad n \in \nsetpos,
$$
where $t_n = \delta n$ for some given observation interval $\delta > 0$, $X_n = X_{t_n}$, $(\eta_n)_{n \in \nsetpos}$ are mutually independent and standard normally distributed noise variables, $\precpar \in \precparsp \eqdef [0, 1]$ is a parameter, and $\varphi$ is a bounded measurable function on $\rset$. We let $\hk_\delta$ denote the (Gaussian) Markov transition kernel of the time-discrete chain $(X_n)_{n \in \nset}$.

@@

In order to illustrate further the $\mathcal{O}(n \precpar)$ bound provided by Theorem~\ref{thm:bias:bound} as well as the stochastic stability of Algorithm~\ref{alg:ideal:PaRIS} established in Corollary~\ref{cor:variance:bound:ideal:PaRIS}, Kalman smoothing was conducted for $n \in \intvect{1}{50}$ on the basis of a skew model with fixed $\varepsilon = 0.1$. Figure~\ref{fig:varying:n} shows, as expected, a linear increase of the bias with $n$ (red line). In addition, displaying also the errors (green lines) of $60$ independent ideal PaRIS replicates (obtained under the skew model with the same algorithmic parametrisation as previously), the same plot confirms the stochastic stability of the ideal PaRIS algorithm, as the variance does not grow faster than linearly with $n$. Finally, for completeness Figure~\ref{fig:varying:n} reports similar errors (blue lines) when the ideal PaRIS evolves under the dynamics of the true model, and in this case the bias is negligible as expected (whereas the variance is still increasing linearly).

>>>>

In the following we will consider an instance of Example~\ref{eq:durham:gallant}, Case~1, where the latent diffusion $(X_t)_{t > 0}$ is an \emph{Ornstein--Uhlenbeck process} \cite{uhlenbeck1930theory} on $\rset$ parameterised by
$$
\mu(x) = - (x - \theta), \quad \sigma(x) \equiv 1,
$$
where $\theta \in \mathbb{R}$ is a parameter.
This process is assumed to be initialised according to the standard normal distribution. Furthermore, conditionally on $(X_t)_{t > 0}$, observations $(Y_n)_{n \in \nset}$ are generated as
$$
Y_n = (1 - \precpar) \varphi(X_n) + \eta_n, \quad n \in \nsetpos,
$$
where $t_n = \delta n$ for some given observation interval $\delta > 0$, $X_n = X_{t_n}$, $(\eta_n)_{n \in \nsetpos}$ are mutually independent and standard normally distributed noise variables, $\precpar \in \precparsp \eqdef [0, 1]$ is a parameter, and $\varphi$ is a bounded measurable function on $\rset$. We let $\hk_\delta$ denote the (Gaussian) Markov transition kernel of the time-discrete chain $(X_n)_{n \in \nset}$.

@@

In order to further illustrate the $\mathcal{O}(n \precpar)$ bound provided by Theorem~\ref{thm:bias:bound} as well as the stochastic stability of Algorithm~\ref{alg:ideal:PaRIS} established in Corollary~\ref{cor:variance:bound:ideal:PaRIS}, Kalman smoothing was conducted for $n \in \intvect{1}{50}$ on the basis of a skew model with fixed $\varepsilon = 0.1$. Figure~\ref{fig:varying:n} shows, as expected, a linear increase of the bias with $n$ (red line). In addition, displaying also the errors (green lines) of $60$ independent ideal PaRIS replicates (obtained under the skew model with the same algorithmic parametrisation as previously), the same plot confirms the stochastic stability of the ideal PaRIS algorithm, as the variance does not grow faster than linearly with $n$. Finally, for completeness Figure~\ref{fig:varying:n} reports similar errors (blue lines) when the ideal PaRIS evolves under the dynamics of the true model, and in this case the bias is negligible as expected.% (whereas the variance is still increasing linearly).

### 2021-08-23 Bernoulli/glo-2020-pm-PaRIS.tex
Corrected typos

\begin{example}[Durham--Gallant estimator]
\label{eq:durham:gallant}
As an illustration, we return to the state-space model framework discussed in Example~\ref{ex:state-space:models}. Let $\Xset \eqdef \rset^{d_x}$ and $\Yset \eqdef \rset^{d_y}$ be equipped with their respective Borel $\sigma$-fields $\Xfd$ and $\Yfd$, and let $(X_t)_{t > 0}$ be some diffusion process on $\Xset$ driven by the homogeneous stochastic differential equation
\begin{equation} \label{eq:SDE}
\rmd X_t = \mu(X_t) \, \rmd t + \sigma(X_t) \, \rmd W_t, \quad t > 0, 
\end{equation}
where $X_0 = 0$, $(W_t)_{t > 0}$ is $d_x$-dimensional Brownian motion, $b : \Xset \to \Xset$ and $\sigma : \Xset \to \Xset^2$ are twice differentiable with bounded first and second order derivatives. In addition, the matrix $\sigma \sigma^\intercal$ is uniformly non-degenerate. Let $(\mathcal{F}_t)_{t > 0}$ be the natural filtration generated by the process $(X_t)_{t > 0}$. The state sequence $(X_t)_{t > 0}$ is latent but partially observed at discrete time points $(t_n)_{n \in \nsetpos}$ which are assumed to be equally spaced for simplicity, \ie, $t_n = t_1 + \delta (n - 1)$ for all $n$ and some $\delta > 0$. Abusing notations, we denote $X_n \eqdef X_{t_n}$ and let $q_\delta$ be the transition density of $(X_n)_{n \in \nsetpos}$. Denote by $\kernel{Q}_\delta$ the transition kernel induced by $q_\delta$. In general, $q_\delta$ is intractable, which makes the problem of computing online, for a given data stream $(y_n)_{n \in \nsetpos}$ in $\Yset$, the sequence of joint-smoothing distributions \eqref{eq:smooth} in models of this sort very challenging. Still, using the Euler scheme, one may, for small $\delta$, approximate $q_\delta$ by 
$$
\bar{q}_\delta(x_n, x_{n + 1}) \eqdef \phi(x_{n + 1}; x_n + \delta \mu(x_n), \delta \sigma^2(x_n)), 
$$ 
where $\phi(\cdot; m, s^2)$ is the density of the Gaussian distribution with mean $m$ and variance $s^2$. Let $\bar{\kernel{Q}}_\delta$ be the transition kernel induced by $\bar{q}_\delta$. Since the approximation $\bar{q}_\delta$ is poor for $\delta$ not small enough, we may instead, as suggested in \cite{durham:gallant:2002}, pick some finer step size $\precpar \in \precparsp_\delta \eqdef \{ \delta / n : n \in \nsetpos \}$ and estimate the density $q_\delta(x_n, x_{n + 1})$ by $q_\delta \langle \zeta \rangle (x_n, x_{n + 1})$, where $\zeta = (\zeta^i)_{i = 1}^L$ are independent draws from some proposal $r(x_n, x_{n + 1}, z) \, \rmd z$ on $\Xset^2 \times \Xfd^{\delta / \precpar - 1}$ and  
$$
q_\delta \langle z \rangle (x_n, x_{n + 1}) \eqdef \frac{1}{L} \sum_{i = 1}^L \frac{\prod_{k = 1}^{\delta / \precpar} \bar{q}_\precpar(z_{k - 1}^i, z_k^i)}{r(x_n, x_{n + 1}, z^i)},  
$$
with $z^i = (z_1^i, \ldots, z_{\delta / \precpar - 1}^i)$ and, by convention, $z_0^i = x_n$ and $z_{\delta / \precpar}^i = x_{n + 1}$. In \cite{durham:gallant:2002}, $r(x_n, x_{n + 1}, z) \, \rmd z$ is the distribution of a discretised (possibly modified) \emph{Brownian bridge}, \ie, Brownian motion started at $x_n$ and conditioned to terminate at $x_{n + 1}$.

@@

\subsubsection*{Case 2} Alternatively, we may assume that $(Y_n)_{n \in \nsetpos}$ are discrete observations of the solution to the stochastic differential equation 
$$
\rmd Y_t = \tilde{\mu}(X_t, Y_t) \, \rmd t + \tilde{\sigma}(X_t, Y_t) \, \rmd \tilde{W}_t, \quad t > 0, 
$$
where $Y = 0$, $(\tilde{W}_t)_{t > 0}$ is $d_y$-dimensional Brownian motion independent of $(W_t)_{t > 0}$ and $\tilde{\mu} : \Xset \times \Yset \to \Yset$ and $\tilde{\sigma} : \Xset \times \Yset \to \Yset^2$ are known functions which are twice differentiable with bounded first and second order derivatives. In addition, the matrix $\tilde{\sigma} \tilde{\sigma}^\intercal$ is uniformly non-degenerate. Denote by $p_\delta$ the transition density of $(X_t, Y_t)_{t > 0}$. In this case, the joint-smoothing distributions can, for a given data stream $(y_n)_{n \in \nsetpos}$, be expressed as path measures \eqref{eq:FK:path} induced by $\ud{n}(x_n, x_{n + 1}) = p_\delta(x_n, y_n, x_{n + 1}, y_{n + 1})$, $n \in \nset$. Since also $p_\delta$ is generally intractable we subject the bivariate process to Euler discretisation, yielding the approximation 
\begin{multline}
\bar{p}_\delta(x_n, y_n, x_{n + 1}, y_{n + 1}) \\
\eqdef \phi \left( x_{n + 1}, y_{n + 1}; (x_n + \delta \mu(x_n), y_n + \delta \tilde{\mu}(x_n, y_n))^\intercal, \delta \, \mbox{diag}(\sigma^2(x_n), \tilde{\sigma}^2(x_n, y_n)) \right) \label{eq:td:bivariate:process}
\end{multline}
of $p_\delta$. Denote by $\bar{\kernel{P}}_\delta$ the Markov kernel induced by $\bar{p}_\delta$. In the case of sparse observations we may improve the approximation $\bar{p}_\delta$ by picking again some finer step size $\precpar \in \mathcal{E}_\delta$ and computing (by swapping, in~\eqref{eq:ukmod:durham:gallant}, $\bar{q}_\delta$ for $\bar{p}_\delta$ and letting $r(x_n, y_n, x_{n + 1}, y_{n + 1}, z) \, \rmd z$ be the distribution of a discretised, $\rset^{d_x + d_y}$-valued Brownian bridge started in $(x_n, y_n)$ and conditioned to terminate in $(x_{n + 1}, y_{n + 1})$) the Durham--Gallant estimator $p_\delta \langle z \rangle$. In this case, $\ukest{n}{z}(x_n, x_{n + 1}) = p_\delta \langle z \rangle (x_n, y_n, x_{n + 1}, y_{n + 1})$, which yields
\begin{equation} \label{eq:ukmod:durham:gallant:case:2}
\ukmod{n}(x_n, \rmd x_{n + 1}) = \bar{p}_\precpar^{\delta / \precpar}(x_n, y_n, x_{n + 1}, y_{n + 1}) \, \rmd x_{n + 1}, 
\end{equation}
where $\bar{p}_\precpar^{\delta / \precpar}$ denotes the transition density of (the $\delta / \precpar$-skeleton) $\bar{\kernel{P}}_\precpar^{\delta / \precpar}$.  
\end{example}

>>>>

\begin{example}[Durham--Gallant estimator]
\label{eq:durham:gallant}
As an illustration, we return to the state-space model framework discussed in Example~\ref{ex:state-space:models}. Let $\Xset \eqdef \rset^{d_x}$ and $\Yset \eqdef \rset^{d_y}$ be equipped with their respective Borel $\sigma$-fields $\Xfd$ and $\Yfd$, and let $(X_t)_{t > 0}$ be some diffusion process on $\Xset$ driven by the homogeneous stochastic differential equation
\begin{equation} \label{eq:SDE}
\rmd X_t = \mu(X_t) \, \rmd t + \sigma(X_t) \, \rmd W_t, \quad t > 0, 
\end{equation}
where $X_0 = 0$, $(W_t)_{t > 0}$ is $d_x$-dimensional Brownian motion, $b : \Xset \to \Xset$ and $\sigma : \Xset \to \rset^{d_x \times d_x}$ are twice differentiable with bounded first and second order derivatives. In addition, the matrix $\sigma \sigma^\intercal$ is uniformly non-degenerate. Let $(\mathcal{F}_t)_{t > 0}$ be the natural filtration generated by the process $(X_t)_{t > 0}$. The state sequence $(X_t)_{t > 0}$ is latent but partially observed at discrete time points $(t_n)_{n \in \nsetpos}$ which are assumed to be equally spaced for simplicity, \ie, $t_n = t_1 + \delta (n - 1)$ for all $n$ and some $\delta > 0$. Abusing notations, we denote $X_n \eqdef X_{t_n}$ and let $q_\delta$ be the transition density of $(X_n)_{n \in \nsetpos}$. Denote by $\kernel{Q}_\delta$ the transition kernel induced by $q_\delta$. In general, $q_\delta$ is intractable, which makes the problem of computing online, for a given data stream $(y_n)_{n \in \nsetpos}$ in $\Yset$, the sequence of joint-smoothing distributions \eqref{eq:smooth} in models of this sort very challenging. Still, using the Euler scheme, one may, for small $\delta$, approximate $q_\delta$ by 
$$
\bar{q}_\delta(x_n, x_{n + 1}) \eqdef \phi(x_{n + 1}; x_n + \delta \mu(x_n), \delta \sigma^2(x_n)), 
$$ 
where $\phi(\cdot; m, s^2)$ is the density of the Gaussian distribution with mean $m$ and variance $s^2$. Let $\bar{\kernel{Q}}_\delta$ be the transition kernel induced by $\bar{q}_\delta$. Since the approximation $\bar{q}_\delta$ is poor for $\delta$ not small enough, we may instead, as suggested in \cite{durham:gallant:2002}, pick some finer step size $\precpar \in \precparsp_\delta \eqdef \{ \delta / n : n \in \nsetpos \}$ and estimate the density $q_\delta(x_n, x_{n + 1})$ by $q_\delta \langle \zeta \rangle (x_n, x_{n + 1})$, where $\zeta = (\zeta^i)_{i = 1}^L$ are independent draws from some proposal $r(x_n, x_{n + 1}, z) \, \rmd z$ on $\Xset^2 \times \Xfd^{\delta / \precpar - 1}$ and  
$$
q_\delta \langle z \rangle (x_n, x_{n + 1}) \eqdef \frac{1}{L} \sum_{i = 1}^L \frac{\prod_{k = 1}^{\delta / \precpar} \bar{q}_\precpar(z_{k - 1}^i, z_k^i)}{r(x_n, x_{n + 1}, z^i)},  
$$
with $z^i = (z_1^i, \ldots, z_{\delta / \precpar - 1}^i)$ and, by convention, $z_0^i = x_n$ and $z_{\delta / \precpar}^i = x_{n + 1}$. In \cite{durham:gallant:2002}, $r(x_n, x_{n + 1}, z) \, \rmd z$ is the distribution of a discretised (possibly modified) \emph{Brownian bridge}, \ie, Brownian motion started at $x_n$ and conditioned to terminate at $x_{n + 1}$.

@@

\subsubsection*{Case 2} Alternatively, we may assume that $(Y_n)_{n \in \nsetpos}$ are discrete observations of the solution to the stochastic differential equation 
$$
\rmd Y_t = \tilde{\mu}(X_t, Y_t) \, \rmd t + \tilde{\sigma}(X_t, Y_t) \, \rmd \tilde{W}_t, \quad t > 0, 
$$
where $Y = 0$, $(\tilde{W}_t)_{t > 0}$ is $d_y$-dimensional Brownian motion independent of $(W_t)_{t > 0}$ and $\tilde{\mu} : \Xset \times \Yset \to \Yset$ and $\tilde{\sigma} : \Xset \times \Yset \to \rset^{d_y \times d_y}$ are known functions which are twice differentiable with bounded first and second order derivatives. In addition, the matrix $\tilde{\sigma} \tilde{\sigma}^\intercal$ is uniformly non-degenerate. Denote by $p_\delta$ the transition density of $(X_t, Y_t)_{t > 0}$. In this case, the joint-smoothing distributions can, for a given data stream $(y_n)_{n \in \nsetpos}$, be expressed as path measures \eqref{eq:FK:path} induced by $\ud{n}(x_n, x_{n + 1}) = p_\delta(x_n, y_n, x_{n + 1}, y_{n + 1})$, $n \in \nset$. Since also $p_\delta$ is generally intractable we subject the bivariate process to Euler discretisation, yielding the approximation 
\begin{multline}
\bar{p}_\delta(x_n, y_n, x_{n + 1}, y_{n + 1}) \\
\eqdef \phi \left( x_{n + 1}, y_{n + 1}; (x_n + \delta \mu(x_n), y_n + \delta \tilde{\mu}(x_n, y_n))^\intercal, \delta \, \mbox{diag}(\sigma^2(x_n), \tilde{\sigma}^2(x_n, y_n)) \right) \label{eq:td:bivariate:process}
\end{multline}
of $p_\delta$. Denote by $\bar{\kernel{P}}_\delta$ the Markov kernel induced by $\bar{p}_\delta$. In the case of sparse observations we may improve the approximation $\bar{p}_\delta$ by picking again some finer step size $\precpar \in \mathcal{E}_\delta$ and computing (by swapping, in~\eqref{eq:ukmod:durham:gallant}, $\bar{q}_\delta$ for $\bar{p}_\delta$ and letting $r(x_n, y_n, x_{n + 1}, y_{n + 1}, z) \, \rmd z$ be the distribution of a discretised, $\rset^{d_x + d_y}$-valued Brownian bridge started in $(x_n, y_n)$ and conditioned to terminate in $(x_{n + 1}, y_{n + 1})$) the Durham--Gallant estimator $p_\delta \langle z \rangle$. In this case, $\ukest{n}{z}(x_n, x_{n + 1}) = p_\delta \langle z \rangle (x_n, y_n, x_{n + 1}, y_{n + 1})$, which yields
\begin{equation} \label{eq:ukmod:durham:gallant:case:2}
\ukmod{n}(x_n, \rmd x_{n + 1}) = \bar{p}_\precpar^{\delta / \precpar}(x_n, y_n, x_{n + 1}, y_{n + 1}) \, \rmd x_{n + 1}, 
\end{equation}
where $\bar{p}_\precpar^{\delta / \precpar}$ denotes the transition density of (the $\delta / \precpar$-skeleton) $\bar{\kernel{P}}_\precpar^{\delta / \precpar}$.  
\end{example}

### 2021-08-23 Bernoulli/glo-2020-theor-results.tex
Corrected typos

In \cite{olsson:westerborn:2014b}, the authors established strong consistency and asymptotic normality of the PaRIS in the framework of fully dominated general state-space HMMs and the bootstrap particle filter, \ie, in the simple case where $\adjfuncforward{n} \equiv 1$ and $\kissforward{n}{n} \equiv \hd{n}$ for all $n$. In the following we will extend these results to the considerably more general setting comprising models \eqref{eq:def:post} and the pseudo-marginal PaRIS in Algorithm~\ref{alg:pm:PaRIS}. More precisely, we will show that each weighted sample $(\epart{n}{i}, \tstat[i]{n}, \ewght{n}{i})_{i = 1}^N$, $n \in \nset$, produced by Algorithm~\ref{alg:pm:PaRIS} satisfies exponential concentration (Theorem~\ref{cor:hoeffding:tau:marginal}) and asymptotic normality (Theorem~\ref{cor:clt:pseudo:marginal:paris}) with respect to the expected additive functional $h_n$ under the \emph{skew} path model 
$$
\postmod{0:n}(\rmd x_{0:n}) \propto \chi(\rmd x_0) \prod_{m = 0}^{n - 1} \ukmod{m}(x_m, \rmd x_{m + 1}), \quad n \in \nset. 
$$
Even though these results are established along the lines of the corresponding proofs in \cite{olsson:westerborn:2014b}, it is the matter of non-trivial adaptations. As explained in Section~\ref{sec:introduction}, a random-weight particle filter (iterated pseudo-marginal forward sampling) can be viewed as a standard particle filter evolving on an extended state space comprising also the states of the auxiliary variables, and hence its convergence follows from standard SMC convergence results \cite{fearnhead2008particle}. On the contrary, since Algorithm~\ref{alg:pm:PaRIS} involves \emph{two} levels of pseudo marginalisation (with respect to both the forward-sampling and the backward-sampling operations), it cannot be described equivalently as a special instance of the original PaRIS (even in its general form given by Algorithm~\ref{alg:ideal:PaRIS}) operating on an extended space. Thus, there is no free lunch when it concerns the theoretical analysis of this scheme. Furthermore, in the case of fully dominated HMMs and when forward sampling is guided by the bootstrap filter, which was the setting in \cite{olsson:westerborn:2014b}, the conditional distribution of the particles given their ancestors (\ie, the marginal of $\rho_n$ in \eqref{eq:cond:instrumental:mixture} with respect to $x$) coincides, at any time point $n$, with the denominator of the backward kernel. Mathematically, this enables a cancellation that facilitates the analysis significantly. However, this simplification is not possible once the particle dynamics is guided by general proposal kernels, which is necessarily the case in Algorithm~\ref{alg:pm:PaRIS}. This complicates the analysis; see Remark~\ref{rem:non-trivial:extension} in supplement for further details.

@@

Our first theoretical result is the following Hoeffding-type concentration inequality, which also plays a critical role for the derivation of the central limit theorem (CLT) in Theorem~\ref{cor:clt:pseudo:marginal:paris}. For every $n \in \nset$, let $\bmaf{\Xfd^n}$ denote the set of additive functionals \eqref{eq:add:func} such that $\addf{m} \in \bmf{\Xfd_m \tensprod \Xfd_{m + 1}}$ for all $m \in \intvect{0}{n - 1}$.   
\begin{theorem}
\label{cor:hoeffding:tau:marginal}
Assume \hypref[assum:biased:estimate]{assum:bound:filter:pseudomarginal}. Then for every $n \in \nset$, $\precpar \in \precparsp$, $h_n \in \bmaf{\Xfd^n}$, and $\K \in \nsetpos$ there exists $(c_n, d_n) \in \rsetpos^2$ such that for all $\N \in \nsetpos$ and $\epsilon > 0$,
$$
\pP\left(\left| \sum_{i = 1}^\N \frac{\ewght{n}{i}}{\sumwght{n}} \tstat[i]{n} - \postmod{0:n} h_n \right| \geq \epsilon \right) \leq c_n \exp \left( - d_n \N \epsilon^2 \right).
$$
\end{theorem}

@@

Theorem~\ref{cor:clt:pseudo:marginal:paris} is established in Section~\ref{sec:proof:prop:clt:pseudo:marginal:paris} in the supplementary paper. The structure of the proof is adopted from \cite{olsson:westerborn:2017} (however, we remark again that it is, as explained above, the matter of a non-trivial extension), and the CLT in Theorem~\ref{cor:clt:pseudo:marginal:paris} is obtained directly as a corollary of a more general CLT for estimators of form $\sum_{i = 1}^\N \ewght{n}{i} \{ f_n(\epart{n}{i}) \tstat[i]{n} + \ftd{n}(\epart{n}{i}) \} / \sumwght{n}$, where $(f_n, \ftd{n}) \in \bmf{\Xfd_n}^2$; see Theorem~\ref{prop:clt:pseudo:marginal:paris} in the supplement.

@@

Importantly, Theorem~\ref{cor:clt:pseudo:marginal:paris} provides, as another by-product, also a CLT for the ideal PaRIS in Algorithm~\ref{alg:ideal:PaRIS}. Indeed, in the case where every $\ud{n}$ is tractable, we may set 
\begin{equation} \label{eq:ideal:case}
\ukest{n}{z}(x_n, x_{n + 1}) = \ud{n}(x_n, x_{n + 1})
\end{equation} 
for all $z$ and define $\ukdist{n}$ arbitrarily. This implies that the relative variance \eqref{eq:def:ukestvar} is identically zero, which eliminates the last term of \eqref{eq:as:var:decomp}. Thus, in this case the asymptotic variance is given by the first two terms of \eqref{eq:as:var:decomp}, but now induced by the original dynamics $(\uk{n})_{n \in \nset}$ (as \eqref{eq:ideal:case} implies that also $\udmod{n} = \ud{n}$). This result is formulated in the following corollary, where we have defined, for each $n \in \nset$ and $m \in \intvect{0}{n}$, the kernel  
\begin{equation} \label{eq:def:retrok}
    \retrok_{m, n}(x_m', \rmd x_{0:n}) \eqdef \delta_{x_m'}(\rmd x_m) \,  
    \tstat{n}(x_m, \rmd x_{0:m - 1})
    \prod_{\ell = m}^{n - 1} \uk{\ell}(x_\ell, \rmd x_{\ell + 1})
\end{equation}
on $\Xset_m \times \Xfd^n$ as well as the centered version 
$$
\retroknorm_{m, n} h (x_m) \eqdef  \retrok_{m, n}(h - \post{0:n} h)(x_m) 
$$
on the same space.

>>>>

In \cite{olsson:westerborn:2014b}, the authors established strong consistency and asymptotic normality of the PaRIS in the framework of fully dominated general state-space HMMs and the bootstrap particle filter, \ie, in the simple case where $\adjfuncforward{n} \equiv 1$ and $\kissforward{n}{n} \equiv \hd{n}$ for all $n$. In the following we will extend these results to the considerably more general setting comprising models \eqref{eq:def:post} and the pseudo-marginal PaRIS in Algorithm~\ref{alg:pm:PaRIS}. More precisely, we will show that each weighted sample $(\epart{n}{i}, \tstat[i]{n}, \ewght{n}{i})_{i = 1}^N$, $n \in \nset$, produced by Algorithm~\ref{alg:pm:PaRIS} satisfies exponential concentration (Theorem~\ref{cor:hoeffding:tau:marginal}) and asymptotic normality (Theorem~\ref{cor:clt:pseudo:marginal:paris}) with respect to the expected additive functional $h_n$ under the \emph{skew} path model 
$$
\postmod{0:n}(\rmd x_{0:n}) \propto \chi(\rmd x_0) \prod_{m = 0}^{n - 1} \ukmod{m}(x_m, \rmd x_{m + 1}), \quad n \in \nset. 
$$
Even though these results are established along the lines of the corresponding proofs in \cite{olsson:westerborn:2014b}, it is a matter of non-trivial adaptations. As explained in Section~\ref{sec:introduction}, a random-weight particle filter (iterated pseudo-marginal forward sampling) can be viewed as a standard particle filter evolving on an extended state space comprising also the states of the auxiliary variables, and hence its convergence follows from standard SMC convergence results \cite{fearnhead2008particle}. On the contrary, since Algorithm~\ref{alg:pm:PaRIS} involves \emph{two} levels of pseudo marginalisation (with respect to both the forward-sampling and the backward-sampling operations), it cannot be described equivalently as a special instance of the original PaRIS (even in its general form given by Algorithm~\ref{alg:ideal:PaRIS}) operating on an extended space. Thus, there is no free lunch when it concerns the theoretical analysis of this scheme. Furthermore, in the case of fully dominated HMMs and when forward sampling is guided by the bootstrap filter, which was the setting in \cite{olsson:westerborn:2014b}, the conditional distribution of the particles given their ancestors (\ie, the marginal of $\rho_n$ in \eqref{eq:cond:instrumental:mixture} with respect to $x$) coincides, at any time point $n$, with the denominator of the backward kernel. Mathematically, this enables a cancellation that facilitates the analysis significantly. However, this simplification is not possible once the particle dynamics is guided by general proposal kernels, which is necessarily the case in Algorithm~\ref{alg:pm:PaRIS}. This complicates the analysis; see Remark~\ref{rem:non-trivial:extension} in the supplement for further details.

@@

Our first theoretical result is the following Hoeffding-type concentration inequality, which also plays a critical role for the derivation of the central limit theorem (CLT) in Theorem~\ref{cor:clt:pseudo:marginal:paris}. For every $n \in \nset$, let $\bmaf{\Xfd^n}$ denote the set of additive functionals \eqref{eq:add:func} such that $\addf{m} \in \bmf{\Xfd_m \tensprod \Xfd_{m + 1}}$ for all $m \in \intvect{0}{n - 1}$.   
\begin{theorem}
\label{cor:hoeffding:tau:marginal}
Assume \hypref[assum:biased:estimate]{assum:bound:filter:pseudomarginal}. Then for every $n \in \nset$, $\precpar \in \precparsp$, $h_n \in \bmaf{\Xfd^n}$, and $\K \in \nsetpos$ there exists $(c_n, d_n) \in \rsetpos^2$ such that for all $\N \in \nsetpos$ and $\delta > 0$,
$$
\pP\left(\left| \sum_{i = 1}^\N \frac{\ewght{n}{i}}{\sumwght{n}} \tstat[i]{n} - \postmod{0:n} h_n \right| \geq \delta \right) \leq c_n \exp \left( - d_n \N \delta^2 \right).
$$
\end{theorem}

@@

Theorem~\ref{cor:clt:pseudo:marginal:paris} is established in Section~\ref{sec:proof:prop:clt:pseudo:marginal:paris} in the supplementary paper. The structure of the proof is adopted from \cite{olsson:westerborn:2017} (however, we remark again that it is, as explained above, a matter of a non-trivial extension), and the CLT in Theorem~\ref{cor:clt:pseudo:marginal:paris} is obtained directly as a corollary of a more general CLT for estimators of form $\sum_{i = 1}^\N \ewght{n}{i} \{ f_n(\epart{n}{i}) \tstat[i]{n} + \ftd{n}(\epart{n}{i}) \} / \sumwght{n}$, where $(f_n, \ftd{n}) \in \bmf{\Xfd_n}^2$; see Theorem~\ref{prop:clt:pseudo:marginal:paris} in the supplement.

@@

Importantly, Theorem~\ref{cor:clt:pseudo:marginal:paris} also provides, as another by-product, also a CLT for the ideal PaRIS in Algorithm~\ref{alg:ideal:PaRIS}. Indeed, in the case where every $\ud{n}$ is tractable, we may set 
\begin{equation} \label{eq:ideal:case}
\ukest{n}{z}(x_n, x_{n + 1}) = \ud{n}(x_n, x_{n + 1})
\end{equation} 
for all $z$ and define $\ukdist{n}$ arbitrarily. This implies that the relative variance \eqref{eq:def:ukestvar} is identically zero, which eliminates the last term of \eqref{eq:as:var:decomp}. Thus, in this case the asymptotic variance is given by the first two terms of \eqref{eq:as:var:decomp}, but now induced by the original dynamics $(\uk{n})_{n \in \nset}$ (as \eqref{eq:ideal:case} implies that also $\udmod{n} = \ud{n}$). This result is formulated in the following corollary, where we have defined, for each $n \in \nset$ and $m \in \intvect{0}{n}$, the kernel  
\begin{equation} \label{eq:def:retrok}
    \retrok_{m, n}(x_m', \rmd x_{0:n}) \eqdef \delta_{x_m'}(\rmd x_m) \,  
    \tstat{n}(x_m, \rmd x_{0:m - 1})
    \prod_{\ell = m}^{n - 1} \uk{\ell}(x_\ell, \rmd x_{\ell + 1})
\end{equation}
on $\Xset_m \times \Xfd^n$ as well as the centered version 
$$
\retroknorm_{m, n} h (x_m) \eqdef  \retrok_{m, n}(h - \post{0:n} h)(x_m) 
$$
on the same space.
